Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-04-19 paperDistance/paper.tex
paper tweaks

\subsection{Setting up the Inference}
Think about presentation, prioritizing/highlighting prior but logically building it from an full inference perspective and needing a prior, then dive into prior stuff. It's confusing right now jumping between inference and prior building !! 
\subsection{Generating the Prior}

Under these assumptions, we would like to infer the true parallax $\varpi_{\true}$ for each star given its measured parallax $\varpi$ and associated uncertainty $\sigma_{\varpi}$ from \tgas.
The strength of our inference method here lies in how we generate our prior information on the true parallax.
Supplementary to this trigonometric measure of distance, the photometry of stars can improve the distance inference.
The intrinsic temperatures and luminosities of stars tightly correlate leading to the observed colors and absolute magnitudes of stars also tightly correlating, usually visualized as the color--magnitude diagram (\cmd).
The tight correlation of the \cmd\ benefits our inference because the absolute magnitude of a star contains distance information.
We include this distance information in the inference of the true parallax by using the \cmd\ as our prior.
Specifically, we use the \cmd\ generated from the data themselves as our prior.
Instead of assuming a functional form for the spatial distribution of stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar models (\citealt{gaia16}), we build a noise-deconvolved estimate of the \cmd\ from the observed colors and absolute magnitudes of our full set of stars.
This \cmd\ model becomes, for each star---given its observed color and magnitude (and uncertainties on those)---a prior pdf for the distance
to that star.

Because our prior in the \cmd\ space, our inference will include not only the parallax from \gaia, but also the photometry of the stars. For inferring the true parallax, including generating our \cmd\ prior, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$. Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\mathbf{y}$ contains a transformation of the dust corrected absolute magnitude

@@

The transformation of the dust corrected absolute magnitude, Equation \ref{eq:transform}, is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain. Gaussian uncertainties are required to generate our prior using \xd\, but more about this and our dust corrections below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the Greene dust map

@@

If life were simple, our inference would be a 1D problem using Bayes theorem looking something like:
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.

@@

where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallas uncertainty dominates over the $J$ band uncertainty, so we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty. This is to keep our uncertainties Gaussian so we can use \xd\ generate our prior below.
This is the probability of the data vector given the true vector, but we
want to flip it around and infer the true vector and therefore the true parallax, or distance. We do this
analogous to Equation \ref{eq:bayes}.

@@

To generate our prior, we fit the data in the slightly transformed \cmd\ space, Equation \ref{eq:transform}, using \xd\ which deconvolves the data with its uncertainties, which are heterogeneous measurement uncertainties. It generates the underlying distribution from which the uncertain data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the marginalized likelihood of the data, marginalized over the true data.

@@

@@

@@

Some discussion of possible failure modes. Including very rare kinds of stars. Should we add gaussians to prior where we know stars exist that aren't well sampled in Gaia (ie WD)

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.

>>>>

\subsection{Setting up the Inference}
 \emph{Think about presentation, prioritizing/highlighting prior but logically building it from an full inference perspective and needing a prior, then dive into prior stuff. It's confusing right now jumping between inference and prior building !!}

Under these assumptions, we would like to infer the true parallax $\varpi_{\true}$ for each star given its measured parallax $\varpi$ and associated uncertainty $\sigma_{\varpi}$ from \tgas.
The strength of our inference method here lies in how we generate our prior information on the true parallax.
Supplementary to this trigonometric measure of distance, the photometry of stars is also informative of their distances.
The intrinsic temperatures and luminosities of stars tightly correlate leading to the observed colors and absolute magnitudes of stars also tightly correlating, usually visualized as the color--magnitude diagram (\cmd).
The tight correlation of the \cmd\ benefits our inference because the luminosity or absolute magnitude of a star, when compared with its brightness or apparent magnitude, contains distance information.
We include this distance information in the inference of the true parallax by using the \cmd\ as our prior.
Specifically, we use the \cmd\ generated from the data themselves as our prior.
Instead of assuming a functional form for the spatial distribution of stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar models (\citealt{gaia16}), we build a noise-deconvolved estimate of the \cmd\ from the observed colors and absolute magnitudes of our full set of stars.
This \cmd\ model becomes, for each star---given its observed color and magnitude (and uncertainties on those)---a prior pdf for the distance to that star.

Because our prior is the \cmd, our inference will include not only the parallax from \gaia, but also the photometry of the stars. For inferring the true parallax, including generating our \cmd\ prior, we use the $J$ and $K$ band photometry from \tmass, as well as the parallax from \gaia. Our true, latent values are the $(J-K)^{true}_n$ color and the $J$ band absolute magnitude $M^{true}_{J,n}$ for each star, which we encapsulate in the true vector $\mathbf{Y}$. The observed data is the dust corrected color $(J-K)^c_n$, the \gaia\ parallax $\varpi_n$ in units of $[mas]$, and the dust corrected $J$ band magnitude $J^c_n$, which we encapsulate in the data vector $\mathbf{y}$. Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\mathbf{y}$ contains a transformation of the dust corrected absolute magnitude

@@

The transformation of the dust corrected absolute magnitude, Equation \ref{eq:transform}, is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain. Gaussian uncertainties are required to generate our prior using \xd\, but more about this and our dust corrections below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the Greene dust map. So our full terminology is

@@

If life were simple, our inference would be a 1D problem using Bayes theorem and look something like:
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.

@@

where $\mathcal{N}$ is a 2D Gaussian function with mean $\mathbf{Y}$ and variance $V_n$.
Here we have assumed that the parallax uncertainty dominates over the $J$ band uncertainty, so we take $J_n^C$ as a point estimate and only propagate the parallax uncertainty. This is to keep our uncertainties Gaussian so we can use \xd\ generate our prior below.

This likelihood is the probability of the data vector given the true vector, but we
want to flip it around and infer the true vector and therefore the true parallax, or distance. We do this
analogous to Equation \ref{eq:bayes}.

@@

\subsection{Generating the Prior}

To generate our prior, we fit the data in the slightly transformed \cmd\ space, Equation \ref{eq:transform}, using \xd\ which deconvolves the data with its uncertainties, which are heterogeneous measurement uncertainties. It generates the underlying distribution from which the uncertain data were drawn from. It represents the distribution as a mixture of Gaussians which maximizes the marginalized likelihood of the data, marginalized over the true data.

@@

where $\alpha$ encapsulates the parameters of each Gaussian in the mixture of Gaussians that defines our \cmd\ prior.

\subsection{Back to the Inference}

@@

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{deltaSimple.png}
\caption{Other example prior: The same visualization as Figure \ref{fig:delta} of the change in the posterior, but here, if instead of prior information containing the \cmd\, the prior information is an exponentially declining spatial density of stars.  }
\label{fig:otherPrior}
\end{figure}

There are other prior information options when inferring distances to stars. A commonly used prior for inferring distances (cite: Tri, CBJ) is the exponentially declining spatial density of stars. It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth. It is a fairly weak prior and basically sets a smooth scale at which it's challenging for stars to lie at, and also allows for negative parallaxes. It is a fairly broad prior so the median \gaia\ star has an increased posterior variance. It has the largest affect on the very noisy and neegative parallaxes and has the nice property of bringing these noisy or negative measurements to reasonable distances. Given this prior, what does the observed \cmd\ look like?  The \cmd\ is broad and the red clump highly stretched.

@@

Some discussion of possible failure modes. Including very rare kinds of stars. Should we add gaussians to prior where we know stars exist that aren't well sampled in Gaia (ie WD)
Inference with our prior is bad for underrepresented populations in the \cmd. For example, it appears there are only $2$ white dwarfs in \gaia\ DR1. With t

@@

\acknowledgments It is a pleasure to thank
  Ana Bonaca (Harvard),
  Andy Casey (Monash),
  Dustin Lang (Toronto) for cross matched 2MASS-TGAS catalogue
and the attendees at the Stars Group Meeting at the Flatiron Institute
Center for Computational Astrophysics for comments and input.
This project was developed in part at the 2016 \acronym{NYC} Gaia Sprint, hosted
by the Center for Computational Astrophysics at the Simons Foundation
in New York City.
